Hydra Battles and Cie

### 2021-02-08 doc/chapter-primrec.tex
redaction on PrimRec and Ackermann function

\section{Introduction}
The definition of primitive recursive functions we use come
from Russel O'Connors formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}, now hosted in
the \texttt{theories/ordinals/Ackermann} directory.

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetics is 32K lines long and is originally structured in 38 modules.

Thus, we propose a slow exploration of this library, through examples and exercises.

@@

\begin{Coqsrc}
Example extEqual_ex1 : extEqual 2 mult (fun x y =>  y * x + x -x) .
Proof.
  intros x y.
\end{Coqsrc}

@@

\begin{remark}
\label{projFunc-order-of-args}
Beware of the conventions used in the \texttt{primRec} library!
The constructor (\texttt{projFunc $n$ $m$})  is associated with the projection $\pi_{n-m,n}$ and \emph{not}
$\pi_{n, m}$.
For instance, the projection $\pi_{2,5}$ defined by $\pi_{2,5}(a,b,c,d,e)=b$ corresponds to the term
(\texttt{projFunc 5 3 H}), where \texttt{H} is a proof of $3<5$.
 This fact is reported in the comments of \texttt{primRec.v}. We presume that this convention makes it easier to define the evaluation function \texttt{evalProjFunc $n$} (see next sub-section) . Trying the other conention is left as an exercise.
\end{remark}

@@

@@

\begin{enumerate}
\item Prove that for any primitive recursive function $f:\mathbb{N}\rightarrow\mathbb{N}\rightarrow\mathbb{N}$, there exists some natural number $n$ depending on $f$, such that, for any $x$ and $y$, 
$f\,x\,y \leq \texttt{Ack}\,n\,(\textrm{max}\,x\,y)$. Let us say that $f$ is \emph{majorized} [by some \texttt{Ack\,$n$}].
\item Show that \texttt{Ack} fails to satisfy this property.
\end{enumerate}

The second part is the easiest one, let us look at it first, and admit the following lemma.

\begin{Coqsrc}
Lemma PR_2_majorized (f: naryFunc 2)(Hf : isPR 2 f)
  : exists (n:nat), forall x y,  f x y <= Ack n (x + y).
\end{Coqsrc}

Then, assume texttt{Ack} is majorized, and find a contradiction.

  Fact Ack_majorized : 
      exists (n:nat), forall x y,  Ack x y <= Ack n (x + y).

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn :  forall x y,  Ack x y <= Ack n (x + y) .

    Lemma contrad : False.
    Proof.
      generalize R07, R08; intros; lia.
    Qed.

Now it remains to prove Lemma \texttt{PR\_2\_majorized}.
If we look at the inductive definition of primitive recursive functions,\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

@@

Thus the lemma we will have to prove is, for any primitive recursive function of abritrary arity $n$ is:

\[
   \exists q\in\mathbb{N},\, \forall x_1,\dots,x_n, 
      f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
\]

@@

>>>>

\section{Introduction}
The definition of primitive recursive functions we use comes
from Russel O'Connors formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}, now hosted in
the \texttt{theories/ordinals/Ackermann} directory.

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetics is 32K lines long and is originally structured in 38 modules.
Thus, we propose a slow exploration of this library, through examples and exercises. Our additions to the original library 
are stored in a sub-directory \texttt{Ackermann/MoreAck}.

@@

\begin{Coqsrc}
Example extEqual_ex1 : extEqual 2 mult (fun x y =>  y * x + x - x) .
Proof.
  intros x y.
\end{Coqsrc}

@@

\begin{remark}
\label{projFunc-order-of-args}
Beware of the conventions used in the \texttt{primRec} library!
The constructor (\texttt{projFunc $n$ $m$})  is associated with the projection $\pi_{n-m,n}$ and \emph{not}
$\pi_{n, m}$.
For instance, the projection $\pi_{2,5}$ defined by $\pi_{2,5}(a,b,c,d,e)=b$ corresponds to the term
(\texttt{projFunc 5 3 H}), where \texttt{H} is a proof of $3<5$.
 This fact is reported in the comments of \texttt{primRec.v}. We presume that this convention makes it easier to define the evaluation function \texttt{evalProjFunc $n$} (see next sub-section) . Trying the other convention is left as an exercise.
\end{remark}

@@

\subsection{A proof by induction on all primitive recursive functions}

@@

\begin{enumerate}
\item Prove that for any primitive recursive function $f:\mathbb{N}\rightarrow\mathbb{N}\rightarrow\mathbb{N}$, there exists some natural number $n$ depending on $f$, such that, for any $x$ and $y$, 
$f\,x\,y \leq \texttt{Ack}\,n\,(\textrm{max}\,x\,y)$ (we say that $f$ is \emph{majorized}  by \texttt{Ack}).
\item Show that \texttt{Ack} fails to satisfy this property.
\end{enumerate}

First, we prove that any primitive function of two arguments is majorized.
If we look at the inductive definition of primitive recursive functions,\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

@@

For instance, proving a property shared by any primitive recursive functions of arity 2 leads to consider the case where that function is obtained by composition with a function of any 
arity $m$. The same problem happens with primitive 
recursion, where a function of arity $n$ is built out of a function of arity $n+1$ and a function of arity $n-1$.

Thus the lemma we will have to prove is the following one:

  \begin{quote}
    For any $n$, and any primitive recursive function $f$ of  arity $n$, there exists some natural number $q$ such that the following inequality holds:
 \[
  \forall x_1,\dots,x_n, 
      f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
\]
 \end{quote}

@@

First, we extend \texttt{max} to vectors of natural numbers (using the notations of module \texttt{VectorNotations} and some more definitions from 
\href{../theories/html/hydras.Prelude.MoreVectors.html}{Prelude.MoreVectors}). So, \texttt{t\,$A$\,$n$} is the type of vectors of $n$ elements of type $A$, and the constants \texttt{cons}, \texttt{nil}, \texttt{map}, etc., refer to vectors and not to lists. Likewise, the notation \texttt{x::v} is an abbreviation for
\texttt{VectorDef.cons x ?n v}.

\begin{Coqsrc}
Fixpoint max_v {n:nat} : forall (v: Vector.t nat n) , nat :=
  match n as n0 return (Vector.t nat n0 -> nat)
  with
    0 => fun v => 0
  | S p => fun (v : Vector.t nat (S p)) =>
             (max (Vector.hd v) (max_v  (Vector.tl v)))
  end.

Lemma max_v_2 : forall x y,  max_v (x::y::nil) = max x y.

Lemma max_v_lub : forall n (v: t nat n) y,
    (Forall (fun x =>  x <= y) v) -> max_v v <= y.

Lemma max_v_ge : forall n (v: t nat n) y,  In  y  v -> y <= max_v v.
\end{Coqsrc}

We have also to convert any application
$(f\,x_1\,x_2\,\dots\,x_n)$ into an application of a function 
to a single argument: the vector of all the $x_i$s.
This is already done in 
Library~\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

\begin{Coqsrc}
Fixpoint evalList (m : nat) (l : Vector.t nat m) {struct l} :
 naryFunc m -> nat :=
  match l in (Vector.t _ m) return (naryFunc m -> nat) with
  | Vector.nil => fun x : naryFunc 0 => x
  | Vector.cons a n l' => fun x : naryFunc (S n) => evalList n l' (x a)
  end.
\end{Coqsrc}

Indeed, (\texttt{evalList $m$ $v$ $f$}) is the application to the vector $v$ of
an uncurried version of $f$.

In Library\href{../theories/html/hydras.Ackermann.MoreAck.AckNotPR.html}{Ackermann.MoreAck.AckNotPR}, we introduce a lighter notation.

\begin{Coqsrc}
(**  uncurried apply:

[v_apply f (x1::x2:: ... ::xn::nil)]  is [f x1 x2 ... xn] 
 *)

Notation "'v_apply' f v" := (evalList _ v f) ( at level 10, f at level 9).

Example Ex2 : forall (f: naryFunc 2) x y,
    v_apply f (x::y::nil) = f x y.
Proof.   intros; now cbn. Qed.

Example Ex4 : forall (f: naryFunc 4) x y z t,
    v_apply f (x::y::z::t::nil) = f x y z t.
Proof.  intros; now cbn. Qed.
\end{Coqsrc}

We are now able to translate in \gallina{} the notion of majoration:

\begin{Coqsrc}
Definition majorized {n} (f: naryFunc n) (A: naryFunc 2) : Prop :=
  exists (q:nat), forall (v: t nat n),
      v_apply f v <= A q  (max_v v).

Definition majorizedPR {n} (x: PrimRec n) A := 
           majorized (evalPrimRec n x) A.

(** For vectors of functions *)

Definition majorizedS {n m} (fs : Vector.t (naryFunc n) m)
           (A : naryFunc 2):=
  exists N, forall (v: t nat n),
      max_v (map (fun f => v_apply f v) fs) <= A N (max_v v).

Definition majorizedSPR {n m} (x : PrimRecs n m) :=
  majorizedS (evalPrimRecs _ _ x).
\end{Coqsrc}

Now, it remains to prove that any primitive function is majorized by \texttt{Ack}.
The three base cases of construction of primitive recursive function are as follows:

\begin{Coqsrc}
Lemma majorSucc : majorizedPR  succFunc Ack.

Lemma majorZero : majorizedPR  zeroFunc Ack.

Lemma majorProjection (n m:nat)(H: m < n):
  majorizedPR (projFunc n m H) Ack.
\end{Coqsrc}

The rest of the cases are proved within a mutual  induction.

\begin{Coqsrc}
Lemma majorAnyPR:  forall n (x: PrimRec n),  majorizedPR  x Ack.
Proof.
  intros n x; induction x using PrimRec_PrimRecs_ind with
                  (P0 := fun n m y => majorizedSPR  y Ack).
  - apply majorSucc.
  - apply majorZero.
  - apply majorProjection. 
  \end{Coqsrc}

  \begin{Coqsrc}
  - (** function composition *)
\end{Coqsrc}
\begin{Coqanswer}
1 subgoal (ID 265)

  n, m : nat
  g : PrimRecs n m
  x : PrimRec m
  IHx : majorizedSPR g Ack
  IHx0 : majorizedPR x Ack
  ============================
  majorizedPR (composeFunc n m g x) Ack
\end{Coqanswer}

\begin{Coqsrc}
  ...
 - (** primitive recursion *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 265)

  n : nat
  x1 : PrimRec n
  x2 : PrimRec (S (S n))
  IHx1 : majorizedPR x1 Ack
  IHx2 : majorizedPR x2 Ack
  ============================
  majorizedPR (primRecFunc n x1 x2) Ack
\end{Coqanswer}

\begin{Coqsrc}
 ...
-  (** empty list of functions *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 266)

  n : nat
  ============================
  majorizedSPR (PRnil n) Ack
\end{Coqanswer}

\begin{Coqsrc}
  ...
- (** non-empty list of functions *)
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 273)

  n, m : nat
  x : PrimRec n
  p : PrimRecs n m
  IHx : majorizedPR x Ack
  IHx0 : majorizedSPR p Ack
  ============================
  majorizedSPR (PRcons n m x p) Ack
\end{Coqanswer}

\begin{Coqsrc}
 ...
Qed.
\end{Coqsrc}

\subsection{Looking for a contradiction}

The following lemma is just a specialization of \texttt{majorAnyPR} to
binary functions (forgetting vectors, coming back to usual notations).

\begin{Coqsrc}
Lemma majorPR2 (f: naryFunc 2)(Hf : isPR 2 f)
  : exists (n:nat), forall x y,  f x y <= Ack n (max x  y).
\end{Coqsrc}

Then, our impossibility proof is just a sequence of easy small steps.

  Fact Ack_majorized : majorized (Ack: naryFunc 2)  Ack.

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn :  forall x y,  Ack x y <= Ack n (max x  y) .

    Lemma contrad : False.
